---
title: "Did the Minimum Wage Kill Jobs? Replicating Card & Krueger (1994)"
description: "A Difference-in-Differences analysis of fast-food employment in New Jersey and Pennsylvania — replicating the key findings from a landmark study showing that a minimum wage increase did not reduce employment."
author: "Yinkai Niu"
date: "2026-04-22"
categories: [Economics, Difference-in-Differences, R, Causal Inference, MGTA 495]
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
---

## Introduction

In April 1992, New Jersey raised its state minimum wage from \$4.25 to \$5.05 per hour — a 19% increase. The textbook prediction from a competitive labor market model is clear: a wage floor set above the equilibrium rate should reduce employment. Employers facing higher mandatory wages will hire fewer workers.

Card and Krueger (1994) challenged this prediction directly. They surveyed 410 fast-food restaurants in New Jersey and neighboring Pennsylvania in February–March 1992 (before the increase) and again in November–December 1992 (after). Pennsylvania's minimum wage remained at \$4.25 throughout, making it a natural control group: geographically adjacent, economically similar, and unaffected by the policy change.

By comparing the change in employment across the two states — a technique now universally called **difference-in-differences (DiD)** — they identified the causal effect of the minimum wage, netting out any common trends in the fast-food industry. Their headline finding: NJ employment rose by roughly 2.76 FTE workers per store relative to PA, flatly contradicting the competitive model's prediction.

This post replicates three core empirical results from the paper as part of **MGTA 495**: Table 2 (descriptive statistics), Table 3 (the DiD estimate), and Table 4 column (i) (the regression version), and adds a graphical visualization of the DiD that the paper only presents in tables.

### Why DiD Enables Causal Inference

A standard comparison of NJ employment before and after the wage increase would be confounded by any other events happening at the same time (e.g., a national recession, seasonal patterns). DiD solves this by differencing away two layers of confounding:

1. **Time-invariant store heterogeneity** — by looking at the *change* in employment (ΔFTE) rather than its level, any fixed characteristic of a store cancels out.
2. **Common time trends** — by subtracting Pennsylvania's average change from New Jersey's, any shock that hit the entire fast-food industry in 1992 is netted out.

$$
\text{DiD} = (\bar{Y}_{NJ,\text{after}} - \bar{Y}_{NJ,\text{before}}) - (\bar{Y}_{PA,\text{after}} - \bar{Y}_{PA,\text{before}})
$$

What remains is the *differential* change experienced by NJ restaurants relative to PA — the only systematic difference between the two groups being New Jersey's minimum wage increase. Under the **parallel trends assumption**, this differential is the causal effect of the policy.

---

## Data

The data are publicly available from David Card's website at UC Berkeley.

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(broom)
library(sandwich)
library(lmtest)

url     <- "https://davidcard.berkeley.edu/data_sets/njmin.zip"
tmp_zip <- tempfile(fileext = ".zip")
tmp_dir <- file.path(tempdir(), "njmin_ck")
dir.create(tmp_dir, showWarnings = FALSE, recursive = TRUE)
download.file(url, tmp_zip, mode = "wb", quiet = TRUE)
unzip(tmp_zip, exdir = tmp_dir, overwrite = TRUE)
dat_file <- list.files(tmp_dir, pattern = "\\.dat$", full.names = TRUE)[1]

var_names <- c(
  "sheet", "chain", "co_owned", "state",
  "southj", "centralj", "northj", "pa1", "pa2", "shore",
  "ncalls", "empft", "emppt", "nmgrs", "wage_st",
  "inctime", "firstinc", "bonus", "pctaff", "meal",
  "open", "hrsopen", "psoda", "pfry", "pentree",
  "nregs", "nregs11",
  "type2", "status2", "date2",
  "empft2", "emppt2", "nmgrs2", "wage_st2",
  "inctime2", "firstin2", "bonus2", "pctaff2", "meal2",
  "open2r", "hrsopen2", "psoda2", "pfry2", "pentree2",
  "nregs2", "nregs112"
)

ck_raw <- readr::read_table(
  dat_file,
  col_names    = var_names,
  col_types    = readr::cols(.default = readr::col_double()),
  na           = c(".", ""),
  show_col_types = FALSE
)

# Build FTE variables
# Permanently closed stores (status2 == 2) get FTE = 0 in wave 2
# following Card & Krueger's Table 3 Column (i) convention
ck <- ck_raw |>
  as_tibble() |>
  mutate(
    state_name = if_else(state == 1, "NJ", "PA"),
    fte1 = empft + 0.5 * emppt + nmgrs,
    fte2 = case_when(
      status2 == 2               ~ 0,
      !is.na(empft2)             ~ empft2 + 0.5 * emppt2 + nmgrs2,
      TRUE                       ~ NA_real_
    ),
    dfte = fte2 - fte1
  )

# Balanced panel: stores with usable data in both waves
balanced <- ck |> filter(!is.na(fte1), !is.na(fte2))

cat("Full dataset:", nrow(ck), "stores |",
    "Balanced panel:", nrow(balanced), "stores\n")
```

The full dataset has 410 stores (331 in NJ, 79 in PA). After restricting to stores with usable employment data in both survey waves, we have **`r nrow(balanced)` stores**.

**Handling of closed restaurants.** Following Card & Krueger's Table 3 Column (i) specification, permanently closed stores (`status2 == 2`) are assigned FTE = 0 in wave 2. Stores that refused or could not be reached are treated as missing and excluded from the balanced panel.

---

## Replication: Table 2 — Descriptive Statistics

Table 2 in the paper reports means and standard errors for key variables by state, confirming that NJ and PA were comparable before the policy change.

```{r}
#| label: table2

smry <- balanced |>
  group_by(state_name) |>
  summarise(
    fte1_m  = mean(fte1,    na.rm = TRUE),
    fte1_se = sd(fte1,      na.rm = TRUE) / sqrt(n()),
    fte2_m  = mean(fte2,    na.rm = TRUE),
    fte2_se = sd(fte2,      na.rm = TRUE) / sqrt(n()),
    w1_m    = mean(wage_st, na.rm = TRUE),
    w1_se   = sd(wage_st,   na.rm = TRUE) / sqrt(n()),
    w2_m    = mean(wage_st2,na.rm = TRUE),
    w2_se   = sd(wage_st2,  na.rm = TRUE) / sqrt(n()),
    n       = n()
  )

nj_s <- smry |> filter(state_name == "NJ")
pa_s <- smry |> filter(state_name == "PA")

t1 <- t.test(fte1    ~ state_name, data = balanced)
t2 <- t.test(fte2    ~ state_name, data = balanced)
t3 <- t.test(wage_st ~ state_name, data = balanced)
t4 <- t.test(wage_st2~ state_name, data = balanced)

fmt <- function(m, se) sprintf("%.2f (%.2f)", m, se)

tibble(
  Variable = c(
    "FTE employment — wave 1 (before)",
    "FTE employment — wave 2 (after)",
    "Starting wage — wave 1 ($/hr)",
    "Starting wage — wave 2 ($/hr)"
  ),
  NJ = c(
    fmt(nj_s$fte1_m, nj_s$fte1_se),
    fmt(nj_s$fte2_m, nj_s$fte2_se),
    fmt(nj_s$w1_m,   nj_s$w1_se),
    fmt(nj_s$w2_m,   nj_s$w2_se)
  ),
  PA = c(
    fmt(pa_s$fte1_m, pa_s$fte1_se),
    fmt(pa_s$fte2_m, pa_s$fte2_se),
    fmt(pa_s$w1_m,   pa_s$w1_se),
    fmt(pa_s$w2_m,   pa_s$w2_se)
  ),
  `t-stat` = c(
    sprintf("%.2f", abs(t1$statistic)),
    sprintf("%.2f", abs(t2$statistic)),
    sprintf("%.2f", abs(t3$statistic)),
    sprintf("%.2f", abs(t4$statistic))
  )
) |>
  kable(
    col.names = c("Variable", "New Jersey", "Pennsylvania", "*t*-stat"),
    caption   = "Table 2: Means of Key Variables (standard errors in parentheses; balanced panel)",
    align     = c("l", "c", "c", "c")
  )
```

The pre-period NJ and PA starting wages are essentially identical (~\$4.25), confirming both groups were near the binding federal minimum before the experiment. The wave-2 NJ wage jumps sharply, reflecting mandatory compliance with the new \$5.05 floor — confirming the treatment actually occurred.

---

## Replication: Table 3 — The DiD Estimate

Table 3 is the heart of the paper. It presents average FTE employment before and after the minimum wage increase, separately for NJ and PA, and computes the difference-in-differences.

```{r}
#| label: table3

m <- balanced |>
  group_by(state_name) |>
  summarise(
    before    = mean(fte1,  na.rm = TRUE),
    before_se = sd(fte1,    na.rm = TRUE) / sqrt(n()),
    after     = mean(fte2,  na.rm = TRUE),
    after_se  = sd(fte2,    na.rm = TRUE) / sqrt(n()),
    change    = mean(dfte,  na.rm = TRUE),
    change_se = sd(dfte,    na.rm = TRUE) / sqrt(n())
  )

nj_m <- m |> filter(state_name == "NJ")
pa_m <- m |> filter(state_name == "PA")

diff_before    <- nj_m$before - pa_m$before
diff_before_se <- sqrt(nj_m$before_se^2 + pa_m$before_se^2)
diff_after     <- nj_m$after  - pa_m$after
diff_after_se  <- sqrt(nj_m$after_se^2  + pa_m$after_se^2)
did            <- nj_m$change - pa_m$change
did_se         <- sqrt(nj_m$change_se^2 + pa_m$change_se^2)

tibble(
  ` ` = c("FTE employment before", "FTE employment after", "Change in FTE employment"),
  `New Jersey`   = c(fmt(nj_m$before, nj_m$before_se),
                     fmt(nj_m$after,  nj_m$after_se),
                     fmt(nj_m$change, nj_m$change_se)),
  `Pennsylvania` = c(fmt(pa_m$before, pa_m$before_se),
                     fmt(pa_m$after,  pa_m$after_se),
                     fmt(pa_m$change, pa_m$change_se)),
  `NJ − PA`      = c(fmt(diff_before, diff_before_se),
                     fmt(diff_after,  diff_after_se),
                     fmt(did,         did_se))
) |>
  kable(
    caption = "Table 3: Average FTE Employment Before and After the Minimum Wage Increase (standard errors in parentheses; balanced panel)",
    align   = c("l", "c", "c", "c")
  )
```

The key cell is the bottom-right: the **DiD estimate of `r round(did, 2)` FTE workers per store** (paper: 2.76). Reading down the NJ − PA column:

- Before the increase, NJ employed fewer workers than PA (`r round(diff_before, 2)` difference).
- After the increase, the gap narrows (`r round(diff_after, 2)`).
- The change in NJ **exceeded** the change in PA by `r round(did, 2)` FTE workers — far from the negative effect that standard theory predicts.

---

## Replication: Table 4, Column (i) — OLS Regression

The DiD estimate has an exact regression equivalent. Rather than comparing group means, we regress the *change* in FTE employment on a New Jersey indicator:

$$
\Delta\text{FTE}_i = \alpha + \beta \cdot \mathbf{1}[\text{NJ}_i] + \varepsilon_i
$$

The intercept $\alpha$ captures the average employment change in Pennsylvania; $\beta$ is the DiD estimate.

```{r}
#| label: table4

mod <- lm(dfte ~ state, data = balanced)
ct  <- coeftest(mod, vcov = vcovHC(mod, type = "HC1"))

tidy(ct) |>
  transmute(
    Term     = c("Intercept (α — PA average change)",
                 "NJ indicator (β — DiD estimate)"),
    Estimate        = round(estimate,  2),
    `Robust SE (HC1)` = round(std.error, 2),
    `t-stat`        = round(statistic, 2),
    `p-value`       = round(p.value,   3)
  ) |>
  kable(
    caption = paste0("Table 4, Column (i): OLS Regression of Change in FTE Employment.",
                     " N = ", nrow(balanced), " stores. R² = ",
                     round(summary(mod)$r.squared, 3), "."),
    align = c("l", "c", "c", "c", "c")
  )
```

The NJ coefficient of **`r round(coef(mod)[2], 2)`** matches the paper's reported 2.76 (paper SE = 1.07). The result is not statistically significant at conventional levels, but the point estimate is positive — the opposite direction from what the competitive model predicts.

---

## Additional Analysis: Visualizing the DiD

The paper only presents these results in tables. The chart below makes the DiD logic immediately visible.

### Visual Summary of the Treatment and Control Groups

This visual summary helps readers compare the before-and-after employment changes for New Jersey and Pennsylvania before interpreting the DiD estimate.

```{r}
#| label: fig-did
#| fig-cap: "Difference-in-Differences: Average FTE Employment in NJ vs PA, Before and After the 1992 Minimum Wage Increase"
#| fig-width: 8
#| fig-height: 5

plot_data <- balanced |>
  select(state_name, fte1, fte2) |>
  pivot_longer(c(fte1, fte2), names_to = "wave", values_to = "fte") |>
  mutate(
    period = if_else(wave == "fte1", "Before\n(Feb–Mar 1992)", "After\n(Nov–Dec 1992)"),
    period = factor(period, levels = c("Before\n(Feb–Mar 1992)", "After\n(Nov–Dec 1992)"))
  ) |>
  group_by(state_name, period) |>
  summarise(
    mean_fte = mean(fte, na.rm = TRUE),
    se_fte   = sd(fte,   na.rm = TRUE) / sqrt(n()),
    .groups  = "drop"
  )

col_nj <- "#d7191c"
col_pa <- "#2c7bb6"

ggplot(plot_data, aes(x = period, y = mean_fte,
                      color = state_name, group = state_name)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4.5) +
  geom_errorbar(aes(ymin = mean_fte - 1.96 * se_fte,
                    ymax = mean_fte + 1.96 * se_fte),
                width = 0.06, linewidth = 0.9) +
  annotate("text", x = 1.5, y = mean(plot_data$mean_fte) - 1,
           label = paste0("DiD = ", round(did, 2), " FTE"),
           size = 5, fontface = "bold", color = "gray30") +
  scale_color_manual(
    values = c("NJ" = col_nj, "PA" = col_pa),
    labels = c("NJ" = "New Jersey (treated)",
               "PA" = "Pennsylvania (control)"),
    name   = NULL
  ) +
  labs(
    title    = "Effect of Minimum Wage Increase on Fast-Food Employment",
    subtitle = "Card & Krueger (1994) — Difference-in-Differences",
    x        = NULL,
    y        = "Mean FTE Employment",
    caption  = "Error bars = 95% CI. NJ raised minimum wage from $4.25 to $5.05 in April 1992; PA did not."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "top",
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

If the minimum wage reduced employment, we would expect the red NJ line to fall after the policy, while the blue PA line stays flat. Instead, **both lines rise, with NJ rising more** — producing a positive DiD estimate of `r round(did, 2)` FTE workers per store.

---

## Discussion

### Replication Fidelity

| Result | Paper | This Replication |
|--------|:-----:|:----------------:|
| NJ FTE before | 20.44 | `r round(nj_m$before, 2)` |
| PA FTE before | 23.33 | `r round(pa_m$before, 2)` |
| DiD estimate | 2.76 | `r round(did, 2)` |

Point estimates match the paper closely. Minor discrepancies can arise from: (1) **sample definition** — the paper presents several cuts of the data with slightly different inclusion rules; (2) **standard error computation** — HC1 heteroskedasticity-robust standard errors are used here.

### Why DiD Supports a Causal Interpretation

DiD's validity rests on the **parallel trends assumption**: in the absence of the minimum wage increase, NJ and PA employment would have evolved along the same trajectory. Several pieces of evidence support this:

- **Geographic proximity**: NJ and PA share a border and face similar labor market conditions.
- **Pre-period wage parity**: As shown in Table 2, wave-1 starting wages in NJ and PA are nearly identical (~\$4.25), confirming both groups started from the same point.
- **Treatment actually occurred**: NJ wages jumped sharply to \$5.05 after April 1992, while PA was unchanged.

Under parallel trends, the DiD estimate of `r round(did, 2)` FTE workers per store is a credible **causal estimate** — not a correlation. Card and Krueger's provocative finding, that a minimum wage increase did not reduce fast-food employment, has been replicated, debated, and refined over three decades, and the DiD methodology they helped popularize is now the gold standard for policy evaluation with natural experiments.

---

## References

Card, David, and Alan B. Krueger. 1994. "Minimum Wages and Employment: A Case Study of the Fast-Food Industry in New Jersey and Pennsylvania." *American Economic Review* 84(4): 772–793.
