---
title: "A/B Testing a Call to Action"
subtitle: "Simulating Key Ideas from Classical Frequentist Statistics"
author: "Yinkai Niu"
date: today
categories: [Statistics, A/B Testing, Simulation, MGTA 495]
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
---

## Introduction

Imagine you run a website with a newsletter sign-up form on the landing page. You have two possible calls to action (CTAs) and you want to know which one persuades more visitors to subscribe:

- **CTA A**: "Sign up for our newsletter here!"
- **CTA B**: "Stay up to date by signing up!"

To find out which performs better, we run an **A/B test**. Each visitor who arrives at the page is randomly assigned to see one of the two CTAs (we'll use a 50/50 split, but the math works for any random allocation), and we then compare the sign-up rates between the two groups. Because assignment is random, any difference in outcomes can be attributed to the CTA itself rather than to who happened to visit. This makes the A/B test a clean, causal experiment.

In this post, we walk through the full statistical machinery behind an A/B test: how to model the data, why our estimator works (Law of Large Numbers), how to quantify uncertainty (Bootstrap), why normal-based inference is valid (Central Limit Theorem), how to perform a formal hypothesis test, and what can go wrong if we peek at results too early.

---

## The A/B Test as a Statistical Problem

Each visitor either signs up (outcome = 1) or does not (outcome = 0). This binary outcome is naturally modeled as a draw from a **Bernoulli distribution**:

$$X_i^A \sim \text{Bernoulli}(\pi_A), \quad X_i^B \sim \text{Bernoulli}(\pi_B)$$

where $\pi_A$ and $\pi_B$ are the true (unknown) sign-up probabilities for each CTA. Because the CTA a visitor sees may influence their probability of signing up, we allow these parameters to differ across groups.

The quantity we care about is the **difference in sign-up rates**:

$$\theta = \pi_A - \pi_B$$

Since we cannot observe $\pi_A$ or $\pi_B$ directly, we estimate $\theta$ using the **difference in sample proportions**:

$$\hat{\theta} = \bar{X}_A - \bar{X}_B = \hat{\pi}_A - \hat{\pi}_B$$

where $\bar{X}_A$ is the fraction of CTA-A visitors who signed up (and similarly for $\bar{X}_B$). Since each observation is 0 or 1, the sample mean is simply the proportion who signed up. The rest of this post explores why this estimator works, how precise it is, and when it can mislead us.

---

## Simulating Data

In a real A/B test we would not know the true values of $\pi_A$ and $\pi_B$ — that is the whole point of running the experiment. For this exercise, however, we set the true values ourselves so we can study how well our estimator behaves when we know the ground truth.

We set $\pi_A = 0.22$ and $\pi_B = 0.18$, giving a true difference of $\theta = 0.04$. We simulate 1,000 independent visitors assigned to each CTA (2,000 total).

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(gt)

set.seed(42)
pi_A <- 0.22
pi_B <- 0.18
n    <- 1000

draws_A <- rbinom(n, 1, pi_A)
draws_B <- rbinom(n, 1, pi_B)

cat(sprintf("CTA A sign-up rate: %.3f (true π_A = %.2f)\n", mean(draws_A), pi_A))
cat(sprintf("CTA B sign-up rate: %.3f (true π_B = %.2f)\n", mean(draws_B), pi_B))
cat(sprintf("Estimated difference θ̂: %.4f (true θ = %.2f)\n",
            mean(draws_A) - mean(draws_B), pi_A - pi_B))
```

Our single-sample estimate of $\theta$ is already close to the true value of 0.04. But how reliable is this estimate? The sections below explore that question.

---

## The Law of Large Numbers

The **Law of Large Numbers (LLN)** states that as the sample size grows, the sample mean converges to the true population mean. Applied to our setting: as we collect more visitors, $\hat{\pi}_A$ converges to $\pi_A$ and $\hat{\pi}_B$ converges to $\pi_B$, so their difference $\hat{\theta}$ converges to the true $\theta = 0.04$. This is what gives us confidence that our estimator will be accurate when the sample is large enough.

To demonstrate this visually, we compute the element-wise differences between the CTA A and CTA B draws, then plot the cumulative (running) average of those differences.

```{r lln}
#| fig-cap: "The running average of the estimated difference converges to the true value of 0.04 as sample size grows — illustrating the Law of Large Numbers."
#| fig-width: 8
#| fig-height: 4

diffs       <- draws_A - draws_B
running_avg <- cumsum(diffs) / seq_along(diffs)

tibble(n = seq_along(running_avg), avg = running_avg) |>
  ggplot(aes(x = n, y = avg)) +
  geom_line(color = "#2c7bb6", linewidth = 0.7, alpha = 0.8) +
  geom_hline(yintercept = 0.04, linetype = "dashed",
             color = "#d7191c", linewidth = 0.9) +
  annotate("text", x = 850, y = 0.055,
           label = "True θ = 0.04", color = "#d7191c", size = 4) +
  labs(
    title    = "Law of Large Numbers: Running Average of θ̂",
    subtitle = "Difference in sign-up rates (CTA A − CTA B)",
    x        = "Number of observations (per group)",
    y        = "Running average of θ̂"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

At small sample sizes, the running average is noisy and volatile — it can swing well above or below 0.04. As the sample size increases, the estimate settles down and converges toward the true value of 0.04. This is the LLN in action: with enough data, we can trust our estimator to be close to the truth.

---

## Bootstrap Standard Errors

Knowing that our estimator is *accurate* (it converges to the truth) is only half the story. We also need to know how *precise* it is for any given sample size. A point estimate $\hat{\theta}$ is not very useful without some measure of uncertainty around it.

**Bootstrapping** is a resampling method that estimates the variability of a statistic without requiring a closed-form formula. The idea is simple: we pretend our observed sample *is* the population, resample from it with replacement many times, compute the statistic of interest each time, and use the standard deviation of those resampled statistics as our estimate of the standard error.

```{r bootstrap}
set.seed(123)
B         <- 1000
boot_ests <- replicate(B, {
  mean(sample(draws_A, n, replace = TRUE)) -
  mean(sample(draws_B, n, replace = TRUE))
})

boot_se <- sd(boot_ests)

# Analytical SE
p_hat_A  <- mean(draws_A)
p_hat_B  <- mean(draws_B)
analytic_se <- sqrt(p_hat_A * (1 - p_hat_A) / n +
                    p_hat_B * (1 - p_hat_B) / n)

# 95% CI using bootstrap SE
theta_hat <- mean(draws_A) - mean(draws_B)
ci_low  <- theta_hat - 1.96 * boot_se
ci_high <- theta_hat + 1.96 * boot_se

tibble(
  Method   = c("Bootstrap SE", "Analytical SE"),
  `Standard Error` = round(c(boot_se, analytic_se), 5)
) |>
  gt() |>
  tab_header(title = "Bootstrap vs. Analytical Standard Error") |>
  fmt_number(columns = `Standard Error`, decimals = 5)
```

The two standard errors are nearly identical, which is reassuring — bootstrapping recovers the same precision estimate as the closed-form formula without requiring any distributional assumptions. The advantage of the bootstrap is that it works equally well for statistics where no clean formula exists; the cost is just computation, which is cheap.

```{r bootstrap-hist}
#| echo: false
#| fig-cap: "The bootstrap distribution of θ̂. The spread of this distribution is the bootstrap standard error; the dashed lines mark the resulting 95% confidence interval."
#| fig-width: 8
#| fig-height: 4

tibble(theta_star = boot_ests) |>
  ggplot(aes(x = theta_star)) +
  geom_histogram(bins = 40, fill = "#2c7bb6", color = "white", alpha = 0.85) +
  geom_vline(xintercept = c(ci_low, ci_high),
             color = "grey30", linetype = "dashed", linewidth = 0.6) +
  geom_vline(xintercept = theta_hat,
             color = "#1a9850", linetype = "solid", linewidth = 0.9) +
  geom_vline(xintercept = 0,
             color = "#d7191c", linetype = "dotted", linewidth = 0.7) +
  labs(
    title    = "Bootstrap distribution of θ̂",
    subtitle = paste0("1,000 resamples; observed θ̂ in green, 95% CI in dashed grey, zero in red"),
    x        = expression(hat(theta)^"*"),
    y        = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

Using the bootstrapped standard error, our **95% confidence interval** for $\theta$ is:

$$\hat{\theta} \pm 1.96 \times \widehat{SE} = [`r round(ci_low, 4)`,\ `r round(ci_high, 4)`]$$

The right way to interpret this: if we re-ran the whole experiment many, many times and built an interval this way each time, about 95% of those intervals would contain the true $\theta$. Reassuringly, our interval covers the true value of 0.04. Whether or not it also covers 0 — the threshold for "no effect" — is the question the formal hypothesis test in two sections will answer.

---

## The Central Limit Theorem

The **Central Limit Theorem (CLT)** tells us that regardless of the shape of the underlying population distribution, the sampling distribution of the sample mean becomes approximately Normal as the sample size grows. Even though each individual outcome is Bernoulli (0 or 1), the *average* of many such outcomes follows a bell-shaped distribution for large samples.

This justifies using Normal-based inference (confidence intervals, z-tests) for our estimator $\hat{\theta}$.

To see this in action, we simulate the sampling distribution of $\hat{\theta}$ at four different sample sizes: $n \in \{25, 50, 100, 500\}$.

```{r clt}
#| fig-cap: "As sample size increases, the sampling distribution of θ̂ becomes increasingly bell-shaped and concentrated around the true value of 0.04 — illustrating the Central Limit Theorem."
#| fig-width: 9
#| fig-height: 7

sim_theta <- function(n_obs, reps = 1000) {
  replicate(reps, {
    mean(rbinom(n_obs, 1, pi_A)) - mean(rbinom(n_obs, 1, pi_B))
  })
}

set.seed(99)
sample_sizes <- c(25, 50, 100, 500)

clt_data <- map_dfr(sample_sizes, ~{
  tibble(n = .x, theta_hat = sim_theta(.x))
})

clt_data |>
  mutate(label = paste0("n = ", n)) |>
  ggplot(aes(x = theta_hat)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, fill = "#2c7bb6", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0.04, color = "#d7191c",
             linetype = "dashed", linewidth = 0.9) +
  facet_wrap(~factor(label, levels = paste0("n = ", sample_sizes)),
             ncol = 2, scales = "free_y") +
  labs(
    title    = "Central Limit Theorem: Sampling Distribution of θ̂",
    subtitle = "1,000 simulated experiments at each sample size. Red dashed line = true θ = 0.04.",
    x        = expression(hat(theta) == bar(X)[A] - bar(X)[B]),
    y        = "Density"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

At $n = 25$, the distribution of $\hat{\theta}$ is rough and somewhat asymmetric — the small sample size means there is high variability. As $n$ increases to 100 and then 500, the histogram becomes smoother, more symmetric, and increasingly bell-shaped, and the spread narrows considerably. This is the CLT in action: even though individual outcomes are binary (Bernoulli), the average of many such outcomes converges to a Normal distribution.

---

## Hypothesis Testing

Now that we know the sampling distribution of $\hat{\theta}$ is approximately Normal for large samples, we can perform a formal hypothesis test.

We set up:

- **Null hypothesis** $H_0: \theta = 0$ — the two CTAs produce the same sign-up rate.
- **Alternative hypothesis** $H_1: \theta \neq 0$ — the two CTAs perform differently.

The goal is to assess whether our data provide enough evidence to reject $H_0$.

To do this, we standardize $\hat{\theta}$ to form a **test statistic**:

$$z = \frac{\hat{\theta} - 0}{SE(\hat{\theta})}$$

Here is the logic step by step:

1. By the CLT, $\hat{\theta}$ is approximately Normal for large samples. Under $H_0$, the mean of that distribution is 0.
2. Dividing a Normal random variable by its standard deviation yields a standard Normal $N(0,1)$. So under $H_0$, $z \;\dot\sim\; N(0,1)$.
3. We know the distribution of $z$ under the null — which is exactly what we need to compute a p-value: the probability of observing a test statistic at least as extreme as ours if $H_0$ were true.

**A note on "t-test" terminology.** This procedure is often called a t-test, but that label is technically imprecise here. The classical t-test arises when data are drawn from a Normal distribution and the variance must be estimated — the ratio then follows an exact t-distribution. Our data are Bernoulli, not Normal, so there is no exact t-distribution result. The CLT gives us approximate Normality, making this closer to a z-test. In practice, for large samples the t-distribution and the standard Normal are nearly identical, so the distinction rarely matters — but it is important to understand why.

```{r hypothesis-test}
z_stat  <- theta_hat / analytic_se
p_value <- 2 * (1 - pnorm(abs(z_stat)))

tibble(
  Statistic = c("θ̂ (estimated difference)",
                "SE(θ̂) (analytical)",
                "z-statistic",
                "p-value (two-sided)"),
  Value = c(round(theta_hat, 4),
            round(analytic_se, 4),
            round(z_stat, 3),
            round(p_value, 4))
) |>
  gt() |>
  tab_header(title = "Hypothesis Test Results",
             subtitle = "H₀: θ = 0  vs.  H₁: θ ≠ 0")
```

```{r hypothesis-decision}
#| echo: false
decision_text <- if (p_value < 0.05) {
  paste0(
    "With a p-value of ", round(p_value, 4),
    ", which is below the conventional threshold of 0.05, we **reject** the null hypothesis. ",
    "The data provide statistically significant evidence that the two CTAs convert at different rates — ",
    "and the sign of $\\hat\\theta$ tells us CTA A outperforms CTA B."
  )
} else {
  paste0(
    "With a p-value of ", round(p_value, 4),
    ", which is above the conventional threshold of 0.05, we **fail to reject** the null hypothesis. ",
    "Note carefully what this does and does not mean: the data are *consistent with* the two CTAs having the same sign-up rate, ",
    "but that is not the same as proving they are equal. With a true effect of only 4 percentage points, ",
    "our sample of n = 1,000 per group sits right at the edge of being able to detect it reliably — ",
    "this experiment simply did not happen to produce a strong enough signal."
  )
}
```

`r decision_text`

This sensitivity — where a true 4-point effect can land on either side of the 0.05 threshold depending on the random draw — is exactly why **sample size matters**, and why power calculations should be part of designing any A/B test.

---

## The T-Test as a Regression

It turns out that the two-sample test we just ran is mathematically equivalent to a simple linear regression. This connection is useful because the regression framework generalizes naturally to more complex experimental designs — multiple treatment arms, controls, interactions — all while producing the same inference in the simple case.

**Setup:** Stack the CTA A and CTA B data into a single dataset with two columns: an outcome $Y_i$ (1 if the visitor signed up, 0 otherwise) and an indicator $D_i$ (1 if the visitor saw CTA A, 0 if they saw CTA B). Then fit:

$$Y_i = \beta_0 + \beta_1 D_i + \varepsilon_i$$

**Interpretation of coefficients:**

1. $\beta_0$ is the expected outcome when $D_i = 0$ — the mean of the B group, estimating $\pi_B$.
2. $\beta_0 + \beta_1$ is the expected outcome when $D_i = 1$ — the mean of the A group, estimating $\pi_A$.
3. Therefore $\beta_1 = \pi_A - \pi_B = \theta$, and the OLS estimate $\hat{\beta}_1$ is numerically identical to $\hat{\theta} = \bar{X}_A - \bar{X}_B$.

```{r regression}
reg_data <- tibble(
  Y = c(draws_A, draws_B),
  D = c(rep(1, n), rep(0, n))
)

mod <- lm(Y ~ D, data = reg_data)

broom::tidy(mod) |>
  mutate(
    term = c("β₀ (Intercept — CTA B mean)", "β₁ (CTA A − CTA B = θ̂)"),
    across(where(is.numeric), ~round(.x, 4))
  ) |>
  gt() |>
  tab_header(
    title    = "Regression: Y ~ D",
    subtitle = "OLS estimate of the difference in sign-up rates"
  ) |>
  cols_label(
    term      = "Term",
    estimate  = "Estimate",
    std.error = "Std. Error",
    statistic = "t-stat",
    p.value   = "p-value"
  )
```

The coefficient $\hat{\beta}_1$ matches $\hat{\theta}$ from the two-sample approach exactly. The standard error differs slightly because OLS assumes a single pooled variance, while our earlier formula allowed separate variances for each group — but for large samples the difference is negligible.

**Why this equivalence matters in practice.** A/B tests are randomized experiments, so we don't need regression to *control for confounding* — randomization already takes care of that. What regression buys us instead is **flexibility**:

- **More than two arms?** Add more indicator variables (A vs. B vs. C and beyond).
- **Continuous covariates available** (device type, geography, prior visits)? Add them to absorb residual variance and tighten standard errors — even though the point estimate of $\theta$ is already unbiased without them.
- **Heterogeneous effects?** Add an interaction term ($D_i \times \text{segment}_i$) to test whether the treatment works differently in different subgroups.
- **Different functional form?** Swap in logistic regression for a probability outcome.

In the simple two-arm case, regression and the two-sample test give the same answer. In every more complex case, regression keeps working while the two-sample formula doesn't generalize. That's why most modern A/B testing infrastructure is built on top of regression.

---

## The Problem with Peeking

Your boss is impatient. Rather than waiting until all 1,000 visitors have been assigned to each group, she wants to check the results after every 100 visitors per group — peeking at $n = 100, 200, \ldots, 1{,}000$ (10 tests in total). If any test comes back significant at $\alpha = 0.05$, she wants to declare a winner and stop early.

**Why is this a problem?** A single test at $\alpha = 0.05$ has a 5% chance of a false positive (rejecting $H_0$ when it is true). But each additional peek is another opportunity to get a false positive. The more times you test, the higher the probability that *at least one* test exceeds the threshold by chance alone — even when there is no real effect.

We can quantify this with a simulation. We set up a world where the null hypothesis is **true**: $\pi_A = \pi_B = 0.20$, so $\theta = 0$. We then run 10,000 experiments, peeking after every 100 observations per group, and record how often at least one peek produces a significant result.

```{r peeking}
#| cache: true

set.seed(777)
n_sims    <- 10000
peek_ns   <- seq(100, 1000, by = 100)
pi_null   <- 0.20

false_pos <- replicate(n_sims, {
  A <- rbinom(1000, 1, pi_null)
  B <- rbinom(1000, 1, pi_null)
  any(sapply(peek_ns, function(k) {
    p_hat_a <- mean(A[1:k])
    p_hat_b <- mean(B[1:k])
    se_k    <- sqrt(p_hat_a * (1 - p_hat_a) / k +
                    p_hat_b * (1 - p_hat_b) / k)
    z_k     <- (p_hat_a - p_hat_b) / se_k
    2 * (1 - pnorm(abs(z_k))) < 0.05
  }))
})

empirical_fpr <- mean(false_pos)

tibble(
  Approach              = c("Single test (no peeking)", "Peeking (10 tests)"),
  `False Positive Rate` = c("5.0%", paste0(round(empirical_fpr * 100, 1), "%"))
) |>
  gt() |>
  tab_header(
    title    = "The Cost of Peeking",
    subtitle = "Empirical false positive rate under H₀: π_A = π_B = 0.20 (10,000 simulations)"
  )
```

The empirical false positive rate under peeking is approximately **`r round(empirical_fpr * 100, 1)`%** — about `r round(empirical_fpr / 0.05, 1)`× the nominal 5%. In other words, even when CTA A and CTA B are *identical*, peeking gives you roughly a `r round(empirical_fpr * 100)`% chance of incorrectly declaring a winner.

This result highlights a fundamental principle of classical frequentist hypothesis testing: the Type I error rate guarantee of 5% applies to a **single, pre-specified test** — not to a sequence of tests on accumulating data. Some practical implications:

- **Pre-register the sample size.** Decide before the experiment how many observations you'll collect, and stick to it.
- **If you must monitor the experiment as it runs**, use methods designed for it: group-sequential designs, alpha-spending functions, sequential probability ratio tests, or always-valid p-values (mSPRT). These explicitly account for multiple looks and control the overall error rate.
- **Be skeptical of "we stopped early because it was significant" stories.** They are exactly the experiments most likely to be false positives.

---

## Wrap-up

We started with a simple business question — *which CTA converts better?* — and built up the statistical scaffolding needed to turn it into a defensible answer. Each piece earned its place:

- **The Bernoulli model** gave us a clean way to talk about a binary outcome.
- **The Law of Large Numbers** told us that with enough data, our estimate $\hat\theta$ will land near the truth.
- **The bootstrap** gave us a way to put error bars on $\hat\theta$ without leaning on a closed-form formula.
- **The Central Limit Theorem** told us *what shape* those error bars take, which let us...
- **Run a hypothesis test** with a real, calibrated false-positive rate.
- **Re-cast the test as a regression** to make the same machinery extend to richer designs.
- **Quantified the cost of peeking**, the most common way A/B tests go quietly wrong in practice.

The same logic — *model, estimator, sampling distribution, inference* — runs through almost every applied statistics problem. Once you've seen it on a CTA test, you've seen most of it.

---

## References & Further Reading

- Efron, B. (1979). *Bootstrap Methods: Another Look at the Jackknife.* The Annals of Statistics, 7(1), 1–26.
- Wasserman, L. (2004). *All of Statistics: A Concise Course in Statistical Inference.* Springer. (Chapters 6–10 cover the LLN, CLT, bootstrap, and hypothesis testing.)
- Johari, R., Koomen, P., Pekelis, L., & Walsh, D. (2017). *Peeking at A/B Tests: Why It Matters, and What to Do About It.* Proceedings of KDD 2017. — The seminal modern paper on the peeking problem and always-valid inference.
- Kohavi, R., Tang, D., & Xu, Y. (2020). *Trustworthy Online Controlled Experiments: A Practical Guide to A/B Testing.* Cambridge University Press.

This post was completed as part of **MGTA 495** at the Rady School of Management, UC San Diego.
